Structured Synchronous Reactive Programming for Game Development \\ \Large{Case Study: On Rewriting Pingus from C++ to \CEU}

### 2017-05-24 paper.tex
...

TODO: falar do state pattern, citar game.patterns, falar que usa estado explicito

\textbf{Case Study: Detecting double-clicks in the \emph{Armageddon button}}

@@

TODO: Structured programming, ...

\textbf{Case Study: The \emph{Bomber action} animation sequence}

@@

@@

\textbf{Case Study: Advancing Pages in the \emph{Story screen}}

@@

@@

\textbf{Case Study: \emph{Bomber action} \code{draw} and \code{update} dispatching}

@@

@@

\textbf{Case Study: Managing the Pingus Lifecycle}

@@

@@

\textbf{Case Study: Global Keys and the Options Menu}

@@

>>>>

\subsubsection{Case Study: Detecting Double-Clicks in the \emph{Armageddon Button}}

@@

\subsubsection{Case Study: The \emph{Bomber Action} Animation Sequence}

@@

\subsubsection{Summary \& Uses in Pingus}

The structured constructs of \CEU provide some advantages in comparison to 
explicit state machines:
%
\begin{itemize}
\item They encode all states with direct sequential code, eliminating shared
      state variables.
\item They handle all states (and only them) in the same contiguous block,
      improving code encapsulation.
\end{itemize}
%
Object-oriented games also adopt the \emph{state pattern} to model state
machines, with subclasses describing each possible state~\cite{games.patterns}.
However, this approach is not fundamentally different from using \code{switch}
or \code{if} branches for each possible state.

\begin{figure}[t]
\begin{verbatim}
Action       Ceu    C++   Explicit State
---------   ----   ----   -----------------
Bomber        23     50   4 state variables
Bridger       75    100   2 state variables
Drown          6     15   1 state variable
Exiter         7     22   2 state variables
Splashed       6     19   2 state variables
\end{verbatim}
\caption{Pingus actions in \CEU and C++.
\label{tab.actions}
}
\end{figure}

Pingus supports 16 actions in the game.
As Figure~\ref{tab.actions} shows, 5 of them implement at least one state
machine and are considerable smaller in \CEU in terms of LoCs when eliminating
the state variables.
%
Considering the other 11 actions, the reduction in LoCs is negligible.
This asymmetry in the implementation of actions illustrates the gains in
expressiveness when describing state machines in direct style.

%As other examples, detecting mouse dragging in the scenario and in the small
%map to move the game viewport also involves state machines.
%State machines also appear in the \emph{FPS counter} and in UI widgets with
%visual feedback.

Among the 65 implementation files in \CEU, we found 29 cases in 25 files using
structured mechanisms to substitute states machines.
They manifest as \code{await} statements in sequence or in aborting constructs
such as \code{par/or} and \code{watching}.

@@

\subsubsection{Case Study: Advancing Pages in the \emph{Story screen}}

@@

\subsubsection{Summary \& Uses in Pingus}

TODO: case 2

The direct style of \CEU has some advantages in comparison to the 
continuation-passing style:
%
\begin{itemize}
\item It uses structured control flow (i.e., sequences and loops) instead of 
      explicit data structures (e.g., stacks) or continuation variables.
\item The activities are decoupled and do not hold references to one another.
\item A single parent class describes the flow between the activities in a 
      self-contained block of code.
\end{itemize}

Continuation passing typically controls the overall structure of the game,
such as screen transitions in menus and level progressions.
%
\CEU uses the direct style techniques in 5 cases involving screen transitions:
the main menu, the level menu, the level set menu, the world map loop, and
the gameplay loop.
%
It also uses the same technique for the loop that switches the pingu actions
during gameplay.

@@

\subsubsection{Case Study: \emph{Bomber Action} \code{draw} and \code{update} Dispatching}

@@

\subsubsection{Summary \& Uses in Pingus}

TODO: case 2

Passive entities subjected to hierarchies require a dispatching architecture
that makes the reasoning about the program harder:

\begin{itemize}
\item The full dispatching chain may go through dozens of files.
\item The dispatching chain may interleave between classes specific to the game
      and also classes from the game engine (possibly third-party classes).
\end{itemize}

In C++, the update subsystem touches 39 files with around 100 lines of code
just to forward \code{update} methods through the dispatching hierarchy.
For the drawing subsystem, 50 files with around 300 lines of code.
The implementation in C++ also relies on dispatching hierarchy for
\code{resize} callbacks, touching 12 files with around 100 lines of code.
%
Most of this code is eliminated in \CEU since abstractions can react directly
to the environment, not depending on hierarchies spread across multiple files.

Note that dispatching hierarchies cross game engine code, suggesting that most
games use this control-flow pattern heavily.
In the case of the Pingus engine, we rewrote 9 files from C++ to \CEU, reducing
them from 515 to 173 LoCs, mostly due to dispatching code removal.

@@

\subsubsection{Case Study: Managing the Pingus Lifecycle}

@@

\subsubsection{Summary \& Uses in Pingus}

TODO: case 2

Lexical lifespan for static instances and natural termination for dynamic
instances provide some advantages in comparison to lifespan hierarchies through
containers:

\begin{itemize}
\item Lexical scope makes an abstraction lifespan explicit in the source code.
\item The memory for static instances is known at compile time.
\item Natural termination makes an instance innocuous and, hence, susceptible
      to immediate reclamation.
\item Abstraction instances (static or dynamic) never require explicit
      manipulation of pointers/references.
\end{itemize}

All entities in a game have an associated lifespan.

The implementation in \CEU has over 200 static instantiations spread across all
65 files.
For dynamic entities, it defines 23 pools in 10 files, with almost 96
instantiations across 37 files.
Pools are used to hold explosion particles, levels and level sets from files,
gameplay \& worldmap objects, and UI widgets.

@@

\subsubsection{Case Study: Global Keys and the Options Menu}

@@

\subsubsection{Summary \& Uses in Pingus}

TODO: case 2

First-class internal events of \CEU provide some advantages in comparison to 
Boost signals:

\begin{itemize}
\item They use the same convenient syntax of external events (e.g.,
      \code{emit}, \code{await}, \code{every}, etc.).
\item They never create infinite dependency loops.
\item They do not require explicit unbinding.
\end{itemize}

The implementation in \CEU uses 39 events for internal communication defined in
23 files with over 200 invocations of \code{emit} and \code{await} spread in
over 50 files.

Internal events are used for resizing entities, broadcasting update and draw
requests, pausing parts of the game, triggering new pingu actions, signalling
collisions, signalling UI interactions, among many others cases.
